Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-25 numerical.tex
going through numerical section, up to sec 5.3 now

% \subsection{Boundary conditions}
% In numerical calculations a finite vertical domain $z\in[-1,1]\zmax$
% is adopted. For $\Gamma\neq1$ we take $\zmax$ to be slightly less than
% $H$ to avoid the zero-density surface. At $z=\pm\zmax$ we impose
% \begin{align}
%   \delta v_z = 0,
% \end{align}
% for a solid boundary, or 
% \begin{align}
%   \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0,  
% \end{align}
% for a free surface, where $\bm{\xi}$ is the Lagrangian
% displacement.

@@

Unless otherwise stated, our fiducial disk model is a
nearly-vertically isothermal disk with $\Gamma=1.011$, $p=-1.5$,
$q=-1$, $\epsilon=0.1$, $\zmax = 10\epsilon r$, and the perturbation
wavenumber $k_x = 200\pi/r$ (or $\hat{k}=20\pi$).

\subsection{VSI in a neutrally stratified disk}\label{vertiso_pertiso} 
We first test our numerical code, as well as the discussion in
\S\ref{iso_discuss}, by considering nearly-isothermal
perturbations. We set $\gamma=1.011=\Gamma$, and effectively switch
off thermal relaxation by choosing $\beta=10^9$. The vertical domain is
$\zmax = 10\epsilon r$. This setup is very similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).

Fig. \ref{lowfreq_eigen} shows the eigenvalues $\sigma = \omega +
\ii\nu$. This plot is effectively identical to that in
\cite{mcnally14}. The fundamental VSI mode, with one node in $W$, has
the smallest $|\sigma|$ and is shown in Fig. \ref{lowfreq_eigenfunc}.  
Its expected growth rate, calculated from Eq. \ref{simple_growth} with  
$L=1$, and the numerically-obtained value are 
\begin{align*}
  \nu = 0.2606\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{simple_growth})},\\
  \nu = 0.2577 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surpsingly good, given that
Eq. \ref{simple_growth} assumes a thin, nearly-Keplerian disk in the 
low-frequency approximation, and
imposes a different boundary condition to that in the numerical
calculation. The eigenfrequency is not very sensitive to the vertical boundary
condition because the expression for $\sigma$, given by
Eq. \ref{integral_relation1}, involve integrands with $\rho$ as the 
weight function, which rapidly decays away from the midplane. The
eigenfunction $W$ is qualitatively proportional to 
$z$, as expected for the fundamental mode according to explicit
solutions discussed in \S\ref{iso_poly}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of the
    vertical shear instability in a neutrally-stratified,
    nearly-vertically isothermal disk  ($\gamma=\Gamma=1.011$). The
    disk parameters are $q=-1$,  
    $p=-1.5$ and $\epsilon=0.1$, while the perturbation radial
    wavenumber is $k_x=200\pi/r$. This is the set up considered in
    \cite{mcnally14}. \label{lowfreq_eigen}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvector_iso}
  \caption{Eigenfunction of the fundamental VSI,
    corresponding to the bottom-left eigenvalue displayed in
    Fig. \ref{lowfreq_eigen} (smallest $|\sigma|$). The
    solid and dashed lines are $\real W$ and $\imag W$, respectively. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

% We have also solved this problem with in the low-frequency 
% approximation by setting $D\to\kappa^2$ in Eq. \ref{ode_w} (with $Q=W$). We
% obtained very similar results, as expected for modes with
% $|\sigma/\Omega_k|\ll1$.

\subsection{Stabilization of the VSI by a positive vertical entropy
  gradient}
We demonstrate the (strong) stabilizing effect of a  
positive vertical entropy gradient by setting $\gamma=2$.  
Since we expect the perturbations to decay rapidly
away from the midplane (see \S\ref{analytic_adia}), we consider a
smaller domain with $\zmax = 2\epsilon r$. 
Other disk and perturbation parameters are the same as in
\S\ref{vertiso_pertiso}.

% We use the same pseudo-spectral method as before. Here we set
% $\zmax = 2\epsilon r$ and impose a free surface. This boundary
% condition was found to be most convenient for numerical
% implementation. Since we expect the perturbations to decay rapidly
% away from the midplane (see \S\ref{analytic_adia}), we do not expect 
% boundary conditions to play an important role, provided that $\zmax$ is
% sufficiently large.

Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
problem. The growth rates are about an order of magnitude smaller than
the previous case. The fundamental VSI mode is that with
$\mathrm{max}|\nu|$. Its expected and numerically-calculated growth
rates compares well, with 
\begin{align*}
  \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
    Eq. \ref{gam2_growth_rate})},\\
  \nu = 0.03672 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
and clearly show that perturbations are rapidly stabilized away from
the midplane. For the adopted disk and perturbation parameters 
($|q|=1$ and $\hat{k} = 20\pi$), Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$,
corresponding to a characteristic decay length scale of $\simeq
0.15\epsilon r$, which is consistent with that observed in 
Fig. \ref{lowfreq_eigenfunc_adia}.

Coincidentally, this decay lengthscale is rougly the height within which
vertical shear is larger than the buoyancy frequency squared. That is,
\begin{align*}
\left|r\frac{d\Omega^2}{dz}\right|\gtrsim N_z^2
\end{align*}
for
\begin{align*}
  \left|\frac{z}{\epsilon r}\right| \lesssim \frac{\epsilon|q|\gamma}{\gamma-1},
\end{align*}
in a thin, vertically isothermal disk. This evaluates to $z\sim
0.2\epsilon r$ for the current disk model, consistent with
Fig. \ref{lowfreq_eigenfunc_adia}.

Furthermore, the growth rate to should be no larger than the
maximum vertical shear available within this decay lengthscale. Using 
Eq. \ref{max_growth} as a rough guide, 
\begin{align}
  \frac{\nu}{\epsilon\Omega_k}<\frac{\epsilon |q|^2\gamma}{2(\gamma-1)},
\end{align}
or $\nu< 0.1\epsilon\Omega_k$ is expected in our disk model,
consistent with numerical results.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
  \caption{Eigenvalues in for the
    vertical shear instability in a nearly-vertically isothermal disk
    ($\Gamma=1.011$) evolved adiabatically with $\gamma=2$ and
    $\beta=10^9$. The disk
    parameters are $q=-1$, 
    $p=-1.5$ and $\epsilon=0.1$, while the perturbation radial
    wavenumber is $k_x=200\pi/r$. \label{lowfreq_eigen_adia}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
  \caption{Fundamental VSI mode in a nearly-vertically
    isothermal disk ($\Gamma=1.011$) evolved adiabatically
    ($\gamma=2$, $\beta=10^9$). This eigenfunction 
    corresponds to the top-left eigenvalue displayed in 
    Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
    The solid and dashed lines are $\real \delta v_z$ and $\imag
    \delta v_z$, respectively.  
    \label{lowfreq_eigenfunc_adia}
  }
\end{figure}

% We have also solved this problem in the low-frequency approximation
% using Eq. \ref{ode_w}---\ref{ode_Q} with the replacement
% $D\to\kappa^2$. We again obtained very similar results.

>>>>

@@

Unless otherwise stated, our disk model is a nearly-vertically
isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
$\epsilon=0.1$.

%mention mode selection -> focus on fundamental corrugation ->
%simulations eventually dominated by them 
%choice of kx is O(10), since nelson observe kxH \sim 30

\subsection{VSI in a neutrally stratified disk}\label{vertiso_pertiso} 
We first calculate the VSI in a neutrally straitified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma$, and effectively switch off thermal relaxation by
choosing $\beta=10^9$. We set the vertical domain to $\zmax =
10\epsilon r$.  Here, we consider a perturbation wavenumber
$\khat=20\pi$ or $k_x = 200\pi/r$. This setup is very similar to that
adopted in \cite{mcnally14}, who calculated the VSI in vertically
isothermal disks subject to isothermal perturbations
($\Gamma=\gamma=1$).  This allows us to test our numerical code.

Fig. \ref{lowfreq_eigen} shows the eigenvalues $\sigma = \omega +
\ii\nu$. The fundamental VSI mode, with one node in $W$, has
the smallest $|\sigma|$ and is shown in Fig. \ref{lowfreq_eigenfunc}. 
The eigenfunction $W$ is qualitatively proportional to 
$z$ (especially for $z\to0$),  as expected for the fundamental mode
according to explicit solutions developed in \S\ref{iso_poly}.    
These plots are consistent with those in \cite{mcnally14}.

The expected growth rate of the fundamental VSI, calculated from
Eq. \ref{simple_growth} with $L=1$, and the numerically-obtained value
are  
\begin{align*}
  \nu = 0.2606\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{simple_growth})},\\
  \nu = 0.2577 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surpsingly good, given that
Eq. \ref{simple_growth} assumes a thin, nearly-Keplerian disk in the 
low-frequency approximation, and
imposes a different boundary condition to that in the numerical
calculation. The eigenfrequency is not very sensitive to the vertical boundary
condition because the expression for $\sigma$, given by
Eq. \ref{integral_relation1}, involve integrands with $\rho$ as the 
weight function, which rapidly decays away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of the VSI in a neutrally-stratified,
    nearly-vertically isothermal disk  ($\gamma=\Gamma=1.011$). The
    disk parameters are $q=-1$, $p=-1.5$, $\epsilon=0.1$, and the perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvector_iso}
  \caption{Eigenfunction of the fundamental VSI,
    corresponding to the bottom-left eigenvalue displayed in
    Fig. \ref{lowfreq_eigen} (smallest $|\sigma|$). The real
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

\subsection{VSI in a stably stratified disk}
We demonstrate the (strong) stabilizing effect of a  
positive vertical entropy gradient ($N_z^2\geq0$) by setting $\gamma=2$.  
Since we expect the perturbations to decay rapidly
away from the midplane (see \S\ref{analytic_adia}), we consider a
smaller domain with $\zmax = 2\epsilon r$. 
Other disk and perturbation parameters are the same as in
\S\ref{vertiso_pertiso}.

Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
problem. The growth rates are much smaller than
those in the previous section. The fundamental VSI mode is that with
$\mathrm{max}\nu$. Its expected and numerically-calculated growth
rates compares well, with 
\begin{align*}
  \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
    Eq. \ref{gam2_growth_rate})},\\
  \nu = 0.03672 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
and clearly show that perturbations are rapidly stabilized away from
the midplane. For the adopted disk and perturbation parameters,
Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$, corresponding
to a characteristic decay length scale of $\sqrt{2/|\alpha|}\epsilon r\simeq 0.2\epsilon r$,
which is consistent with that observed in  Fig. \ref{lowfreq_eigenfunc_adia}.

This decay lengthscale is rougly the height within which
vertical shear is larger than the buoyancy frequency squared. That is,
\begin{align*}
\left|r\frac{d\Omega^2}{dz}\right|\gtrsim N_z^2
\end{align*}
for
\begin{align*}
  \left|\frac{z}{\epsilon r}\right| \lesssim \frac{\epsilon|q|\gamma}{\gamma-1},
\end{align*}
in a thin, vertically isothermal disk. This evaluates to $|z|\lesssim
0.2\epsilon r$ for the current disk model, consistent with
Fig. \ref{lowfreq_eigenfunc_adia}.

Furthermore, the growth rate to should be no larger than the
maximum vertical shear available within this decay lengthscale. Using 
Eq. \ref{max_growth} as a rough guide, 
\begin{align}
  \frac{\nu}{\epsilon\Omega_k}<\frac{\epsilon |q|^2\gamma}{2(\gamma-1)},
\end{align}
or $\nu< 0.1\epsilon\Omega_k$ in our disk model,
consistent with numerical results.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
  \caption{Eigenvalues in for the
    VSI in a nearly vertically isothermal disk
    ($\Gamma=1.011$) evolved adiabatically with $\gamma=2$. The disk
    parameters are $q=-1$, 
    $p=-1.5$, $\epsilon=0.1$, and the perturbation radial
    wavenumber is $\khat=20\pi$. \label{lowfreq_eigen_adia}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
  \caption{Fundamental VSI mode in a nearly vertically
    isothermal disk ($\Gamma=1.011$) evolved adiabatically
    with $\gamma=2$. This eigenfunction 
    corresponds to the top-left eigenvalue displayed in 
    Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
    The  real (imaginary) part of $\delta v_z$ are shown as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc_adia}
  }
\end{figure}

Although a positive vertical entropy gradient is stabilizing, we note
that $N_z^2\propto z^2$  while vertical shear $rd\Omega^2/dz\propto
z$, so that there is always a region sufficiently close to the
midplane in which vertical shear dominates over bouyancy and the VSI
can operate. However, the growth rates are small and the instability
only affects the regions close the midplane. These modes may not be
important in practice.

### 2015-01-25 setup.tex
adding youdin notes
almost done with numerical results section
going through numerical section, up to sec 5.3 now

\section{Baroclinic disk equilibra}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. 
We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively.

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)}
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant. 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
and we are interested in neutrally or sub-adiabatically stratified
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

The equilibrium disk satisfies
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r},\label{hori_equilibrium} 
\end{align} 
where $\Phi_* = -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational
potential of the central star. Note that the disk is in general
baroclinic with $\Omega = \Omega(r,z)$, since
Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} imply, with
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p z}\frac{dK}{dr}.\label{vertical_shear}
\end{align}

The explicit expression for the density field for
$\Gamma\neq1$ is given  by
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right),
\end{align}
Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
$\rho(r,H)=0$. If we parametrize
\begin{align}
  H(r)\equiv h r,
\end{align}
then the imposed sound-speed and disk thickness are related by 
\begin{align}
  h^2(r) = \left[1-\frac{\epsilon^2(r)}{\Gamma-1}\right]^{-2}-1. 
\end{align}

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define the radial shear rate  
\begin{align}
  S(r,z) \equiv r\frac{\p\Omega}{\p r},  
\end{align}
and the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv 2\Omega(2\Omega + S). 
\end{align}

@@

\subsection{Vertically isothermal disks}
For the special case $\Gamma=1$ the disk is isothermal in the
vertical direction. The equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ is then 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. The vertically isothermal disk has no
surface, but we can define a characteristic scale-height 
\begin{align}
  \Hiso \equiv \epsilon r.% \quad\quad(\Gamma \equiv 1),   
\end{align} 
and in this case the characteristic aspect-ratio $h = \epsilon$.

@@

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), 
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation
the ratio of specific heats $\gamma$ is assumed constant, 
and we have included a thermal relaxation term with timescale
$t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
parameter.

>>>>

\section{Baroclinic disk equilibra}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. 
We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. The equilibrium disk satisfies
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r},\label{hori_equilibrium} 
\end{align} 
where $\Phi_* = -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational
potential of the central star.

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.  
% From $\epsilon$ we can define a
% characteristic scale-height  
% \begin{align}
%   \Hiso \equiv \epsilon r.% \quad\quad(\Gamma \equiv 1),   
% \end{align} 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
and we are interested in neutrally or sub-adiabatically stratified
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

The explicit expression for the density field for
$\Gamma\neq1$ is given  by
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right),
\end{align}
Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
$\rho(r,H)=0$. If we parametrize
\begin{align}
  H(r)\equiv h r,
\end{align}
then the imposed sound-speed and disk thickness are related by 
\begin{align}
  h^2(r) = \left[1-\frac{\epsilon^2(r)}{\Gamma-1}\right]^{-2}-1. 
\end{align}
Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define the radial shear rate  
\begin{align}
  S(r,z) \equiv r\frac{\p\Omega}{\p r},  
\end{align}
and the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv 2\Omega(2\Omega + S). 
\end{align}

Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot}, but is also
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr}.\label{vertical_shear}
\end{align}

@@

\subsection{Vertically isothermal disks}
For the special case $\Gamma=1$ the disk is isothermal in the
vertical direction. The equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ is then 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks for simplicitly.

@@

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), \label{energy_eq}
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation
the ratio of specific heats $\gamma$ is assumed constant, 
and we have included a thermal relaxation term with timescale
$t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
parameter.
